---
title: "Lab 5: Import and Export"
subtitle: "Stat 33B"
author: "Prof. Gaston Sanchez"
output: 
  html_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
```

> ### Learning Objectives
>
> -   Get to know various options to export data
> -   Get to know various options to import data tables in R
> -   Practice using reading tables functions from package `"utils"`

### General Instructions

-   Write your descriptions, explanations, and code in an `Rmd` (R
    markdown) file.
-   Name this file as `lab05-first-last.Rmd`, where `first` and `last`
    are your first and last names (e.g. `lab05-gaston-sanchez.Rmd`).
-   Knit your `Rmd` file as an html document (default option).

------------------------------------------------------------------------

## 1) Filestructure

In your computer, create a directory (i.e. a folder) `lab05`. Likewise,
create four subdirectories---`data`, `code`, `output`, `images`---as
depicted in the file-structure shown below. You can try to use the
`dir.create()` to create such directories (no need to include code
chunks for this step in your Rmd file). Also, make sure to place your
`Rmd` file inside the `lab05` directory.

        lab05/
          lab05-first-last.Rmd
          data/
          code/
          output/
          images/

```{r}
dir.create("data")
dir.create("code")
dir.create("output")
dir.create("images")
```

### 1.1) Your Turn: using `write.csv()`

a)  Use the function `write.csv()` to export (or save) the matrix
    `state.x77` to a data file `states.csv` in the `data/` directory.
    Also, see how to use the argument `row.names` to avoid including a
    first column of numbers.

```{r}
# your code
write.csv(state.x77,file = "data/states.csv",row.names = FALSE)
```

------------------------------------------------------------------------

## 2) Diverting R output to external files

Say you are interested in exporting the `summary()` statistics of the
column `Area` in `state.x77`. Specifically, say you want to export the
numeric output exactly in the same way they are displayed by R (see
output below):

```{r}
summary(state.x77[ ,"Area"])
```

One naive option would be to manually copy the text displayed on the
console, and then paste it to a text file. While this may work, it is
labor intensive, error prone, and highly irreproducible. A better way to
achieve this task is with the `sink()` function. Here's how:

``` r
# divert output to text file inside folder output/
sink(file = 'output/summary-area.txt')
summary(state.x77[ ,"Area"])
sink()
```

The first call to `sink()` opens a connection to the specified file, and
then all outputs are diverted to that location. The second call to
`sink()`, i.e. the one without any arguments, closes the connection.

### 2.1) Your turn: using `sink()`

a)  Export the output of `summary()` by selecting columns `Population`,
    `Income`, and `Area` of `state.x77` to a text file
    `states-summary.txt`, in the `output/` folder.

```{r}
# your code
sink("output/states-summary.txt")
summary(state.x77[,c("Population","Income","Area")])
sink()
```

<br>

b)  Export the output of `summary()` by selecting rows of `state.x77`
    with `Area` greater than 100,000 square miles, and columns
    `Illiteracy`, `Murder` and `Area`. Export this `summary()` to a text
    file `summary-large-states.txt` in the `output/` folder.

```{r}
# your code
sink("output/summary-large-states.txt")
summary(state.x77["Area">100000,c("Illiteracy","Murder","Area")])
sink()
```

------------------------------------------------------------------------

## 3) Exporting some graphics

In the same way that R output as it appears on the console can be
exported to some files, you can do the same with graphics and plots.
Actually, saving plot images is much more common than `sink()`ing
output.

Base R provides a wide array of functions to save images in most common
formats:

-   `png()`
-   `jpeg()`
-   `tiff()`
-   `bmp()`
-   `svg()`
-   `pdf()`

Like the writing table functions such as `write.table()` or
`write.csv()`, and the `sink()` function, the graphics device functions
require a file name to be provided. Here's how to save a histogram of
`Income` in png format to the file `histogram-income.png` inside the
folder `images/`:

``` r
# saving a scatterplot in png format
png(filename = "images/histogram-income.png")
hist(state.x77[ ,"Income"])
dev.off()
```

-   The function `png()` tells R to save the image in PNG format, using
    the provided filename.

-   Invoking `png()` will open a graphics device; not the graphics
    device of RStudio, so you won't be able to see the graphic.

-   The `plot()` function produces the scatterplot.

-   The function `dev.off()` closes the graphics device.

<br>

### 3.1) Your turn: using `png()`, etc

a)  Use `png()` to save a scatterplot of `Illiteracy` (x-axis) and
    `Murder` (y-axis) with `plot()`. Save the graph in the `images/`
    folder, naming the file as `scatterplot-illiteracy-murder.png`.

```{r}
# your code
png("images/murder.png")
plot(state.x77[,"Illiteracy"],state.x77[,"Murder"],xlab = "Illiteracy",ylab = "Murder")
dev.off()
```

<br>

b)  Use `pdf()` to save the previous histogram of `Income` in PDF
    format, with dimensions (width x height) 6 x 4 inches.

```{r}
# your code
pdf("images/histogram-income.pdf",width = 6, height = 4)
hist(state.x77[ ,"Income"])
dev.off()
```

------------------------------------------------------------------------

## 4) Importing Data Tables

Data sets come in many different presentations. One common format is
that of a data table---like a spreadsheet---although you can find data
in other formats (especially when the data is in its raw version).

The folder for this lab (in bCourses) contains the following data
text-files:

-   `starwars1.txt`: *space separated* file

-   `starwars2.csv`: *comma separated value* (CSV) file

-   `starwars3.txt`: file with both metadata and data

-   `starwars4.dat`: *fixed width format* (fwf) file

**Download these files and place them in the same directory of your Rmd
file**

<br>

Because data tables are so ubiquitous, it's good if you learn how to
import them in R. In this section we are going to review various aspects
that have to do with reading in tables in R.

We will talk about the following functions (and packages)

-   `read.table()`
-   `read.csv()`
-   `read.delim()`
-   `read.fwf()`

### 4.1) Function `read.table()`

`read.table()` is the workhorse function to import tables in R. The
mandatory input is the file-path of the file to be imported. The rest of
the arguments have default values. Some arguments that are commonly
specified are:

-   `header`: whether the file contains column names

-   `sep`: field separator character

-   `col.names`: vector of optional names for the columns

-   `na.strings`: character vector to indicate how missing values are
    encoded

-   `colClasses`: character vector to indicate data-types for the
    columns

-   `nrows`: maximum number of rows to read in

-   `skip`: number of lines to skip before beginning to read data

Here's an example to import the data in `starwars1.txt` with
`read.table()`

```{r}
# reading data from your working directory
sw1 <- read.table(
  file = "starwars1.txt", 
  header = TRUE)

sw1
```

Notice that R uses some heuristics to decide---with an educated
guess---the most appropriate data-type for each column. You can take
full control of this decision with the `colClasses` argument. For
example, say you want to import columns `gender` and `jedi` as
`"factor"`:

```{r}
# reading data from your working directory
sw1 <- read.table(
  file = "starwars1.txt", 
  header = TRUE,
  colClasses = c(
    "character", # name
    "factor",    # gender
    "double",    # height
    "integer",   # weight
    "factor",    # jedi
    "character", # species
    "character"  # weapon
  )
)

sw1
```

<br>

### 4.2) Your Turn: CSV file

See the documentation of `read.csv()`, and use this function to import
the data in `starwars2.csv`. Missing values in this file are encoded
with a question mark character `"?"`. Name this data as `sw2`, and
display it.

```{r}
# your code
sw2 <- read.csv(file = "starwars2.csv",na.strings = "?")
sw2
```

<br>

### 4.3) Your Turn: importing a few columns

Keep looking at the documentation of `read.table()`, and focus on the
argument `colClasses`. How would you specify a vector for this argument
in order to import just the columns `name`, `height`, and `weapon`? Name
this data as `sw2b`, and display it.

```{r}
# your code
sw2b <- read.csv(file = "starwars2.csv",colClasses = c("character","NULL","double","NULL","NULL","NULL","character"))
sw2b
```

<br>

### 4.4) Your Turn: File with metadata

Take a look at the file `starwars3.txt`. As you can tell, this file has
three sections:

-   The first four lines of text have some general information with the
    description of the file, its format, and units of measurement for
    `height` and `weight`

-   The next section, from line 6 to 13, indicates the names of the
    columns

-   The last section, from line 15 to 23, have the actual data table

Your goal is to write commands, without hardcoding any values, in order
to import the data table (with their corresponding column names). We
recommend the following strategy:

1)  use `readLines()` to import the first 13 lines into a character
    vector

2)  from the character vector of step (1), extract those elements that
    correspond to the column names; this will give you a character
    vector with just the column names

3)  use a reading table function to import the data values from line 16
    till the end of the file. *Hint*: argument `skip`---and possibly
    `nrows`---is your friend. Likewise, use the vector of column names
    previously extracted in the step (2) to specify values for the
    argument `col.names`.

4)  Name this data as `sw3`, and display it.

```{r}
# your code
str <- readLines(con = "starwars3.txt", n = 13)
str = str[7:13]
sw3 <- read.table("starwars3.txt",skip = 15,col.names = str)
sw3
```

<br>

### 4.5) Your Turn: Fixed Width Format

See the documentation of the function `read.fwf()`, and use this
function to import the data in `starwars4.dat`.

One suggestion to complete this operation is to specify arguments:

-   name of the `file`

-   skip the first line

-   specify `col.names` with a character vector using the names in the
    first line of the file: e.g. `name`, `gender`, `height`, `weight`

-   specify column widths with a numeric vector; the widths are as
    follow: 16 (1st column), 7 (2nd col), 7 (3rd col), and 7 (4th col)

-   name this file `sw4` and display it

```{r}
# your code
sw4 <- read.fwf("starwars4.dat",skip = 1, col.names = c("name","gender","height","weight"),widths = c(16,7,7,7))
sw4
```
